\textsf{jaxoplanet}: Hardware-Accelerated Computation of Stellar Light Curves

### 2024-10-16 latex/paper.tex
feat: speed benchmark done on realistic examples

In \autoref{fig:precision_limbdark}, we compare the precision of our limb darkened light curve model with \textsf{exoplanet} Python package (which uses the same formalism as \cite{Agol2020} but limited to linear and quadratic limb darkening).from shows the relative error in the flux of a limb-darkened star occulted by an opaque body for increasing orders of a polynomial limb darkening law. In this figure, we also report the error of the flux computed using the \textsf{exoplanet} Python package (which uses the same formalism as \cite{Agol2020} but limited to linear and quadratic limb darkening).

@@

A set of unitary polynomial limb darkening coefficients of order 20 leads to spherical harmonics coefficients as large as $10^6$. This explains the larger errors observed in \autoref{fig:precision_limbdark} (error versus limb darkening order) compared to the ones shown in \autoref{fig:precision_degree} (error versus degree of the spherical harmonics). We note that the errors scale with the values of $\bvec{u}$.\\\\
With these results, we validate the precision of \textsf{jaxoplanet} and show it is on per with the legacy C++ \textsf{starry} implentation\footnote{version 1.2.0 \citep{starry_120}} described in \cite{starry}.\\\\
As described in \autoref{solution_vector}, we compute occultation light curves using the Gauss Legendre quadrature to approximate the $\mathcal{P}$ integral involved in the expression of the solution vector $\bvec{s}^\mathsf{T}$. Hence, the precision of the computed flux $f$ depends on the order of the Legendre polynomial which defines the number of points used to approximate $\mathcal{P}$. For this reason, users must carefully set the order parameter $q$ to reach the precision required for their application, as shown in \autoref{fig:relative_error_1}.\\\\

@@

The minimum error is always achieved for higher values of $q$, which comes with increased computation time (see \autoref{speed}). Hence users must carefully set the order parameter to balance precision and computational speed. As the minimum error reached also depends on the degree of the spherical harmonics basis used, \autoref{fig:precision_order_gausslegendre_degree} shows the errors on the flux for each degree of the spherical harmonics basis, for an occultor with relative radii $r=0.01$, $r=1$ and $r=100$.
        \includegraphics[width=\textwidth]{../workflows/precision/figures/error_order_degree.pdf}
        \caption{Maximum errors in the overall flux $f$ for all degrees of the spherical harmonics components for different occultor radii and orders $q$ of the Gauss-Legendre quadrature used to compute the P integral (see \autoref{solution_vector}). Errors shown in this figure correspond to the maximum values of the scaled errors shown in \autoref{fig:precision_s}, maxima taken over the range of values $b$.}
        \label{fig:precision_order_gausslegendre_degree}

@@

One of the advantage of the \textsf{JAX} library is its ability to perform hardware-accelerated computations on CPUs and GPUs. In this section, we evaluate the speed of the \textsf{jaxoplanet} implementation and compare it with the C++ implementation of \textsf{starry}. For quadratically limb darkened light curves, we also compare our implementation with the one provided by the \textsf{exoplanet} Python package.\\\\
In \autoref{fig:speed_order_gausslegendre}, we show the evolution of the time required to compute the occultation light curve of a limb darkened star and that of a non-uniform star described by spherical harmonics of degree $l_{max}=20$, depending on the order $q$ of the numerical integration. On CPU, our benchmark is done on the single core of an Apple M2 max chip, while on GPU we use the single core of an NVIDIA Tesla V100 processor.\\\\
\begin{figure}[H]
    \begin{center}
        \includegraphics[width=\textwidth]{../workflows/speed/figures/speed_vs_order.pdf}
        \caption{Computation time of the occultation light curve of a limb-darkened star (left) and a non-uniform star described by spherical harmonics up to order $l_{max}=20$ (right) depending on the order $q$ of the Gauss-Legendre quadrature used to compute the $\mathcal{P}$ integral numerically (see \autoref{solution_vector}). The processing time reported for \textsf{exoplanet} and \textsf{starry} is independent of $q$ as these two codes provide closed-form solutions for the integral $\mathcal{P}$. Light curves are computed for $10\,000$ points in transit and an occultor radius $r=0.1$.}
        \label{fig:speed_order_gausslegendre}
    \end{center}
\end{figure}
Finally, \autoref{fig:speed_N} shows the evaluation time of \textsf{jaxoplanet} as a function of the number of points in the light curve compared to \textsf{exoplanet}, for a quadratically limb darkened star, and \textsf{starry}, for a surface map of degree $l_{max} = 20$.
\begin{figure}[H]
    \begin{center}
        \includegraphics[width=\textwidth]{../workflows/speed/figures/speed_vs_N.pdf}
        \caption{Computation time of the occultation light curve of a limb-darkened star (left) and a non-uniform star described by spherical harmonics up to order $l_{max}=20$ (right) depending on the number of points in the light curve. Light curves are computed at $q=200$ and an occultor radius $r=0.1$.}
        \label{fig:speed_N}
    \end{center}
\end{figure}
These benchmarks are produced for values of $q$ which guarantee that our models reach the same precision as state-of-the-art codes. As discussed in \autoref{discussion}, \textsf{jaxoplanet} can be further optimized by reducing the order $q$ of numerical integration while keeping the precision required for specific applications. In this context, the \textsf{jaxoplanet} Python package provide helper functions to estimate the precision of the light curve model given the relative radius $r$, the spherical harmonics degree $l_{max}$ (if general non-uniform maps are requires), the order of the polynomial limb darkening, and the order $q$ of the numerical integration.\\\\

\section{Discussion}\label{discussion}

In \autoref{precision}, we showed that \textsf{jaxoplanet} is on par with the C++ implementation of \textsf{starry} in terms of precision. As relatively high values of $q$ are required to achieve this performance, results from \autoref{speed} are deceiving and suggest that \textsf{jaxoplanet} is about an order of magnitude slower than \textsf{exoplanet} or \textsf{starry} on CPU, and only advantageous on GPU for very large datasets. Although we expect to further optimize our \textsf{JAX} implementation and bridge this gap in the future, \textsf{jaxoplanet} can currently outperform state-of-the-art codes for most realistic datasets. Indeed, models computed at low values of $q$, which are very fast, match the precision of measurements from modern instruments.\\\\
To demonstrate our point, we first consider the transmission spectroscopy of TRAPPIST-1 b performed with the JWST/NIRISS instrument (GO-2589, \citealt{Lim2023}). These observations reached an estimated precision of 89 ppm for a planet with a relative radius of the order $r\sim0.1$. In this context, numerical integration with an order as low as $q=10$ guarantees that our model reaches a precision of $10^{-16}$, an execution time on par with \textsf{exoplanet} on CPU and up to an order of magnitude faster on GPU for large datasets (see left plot of \autoref{fig:real_benchmark}). The second example we consider is the secondary ecclipse mapping of HD 189733b ($r\sim0.3$) using Spitzer's 8 micron channel \citep{majeau2012} obtained at a precision of about 1 ppt. For a spherical harmonics map of degree $l_{max}=3$ (although $l_{max}=1$ is enough in this case), $q=10$ guarantees a model with a precision of $10^{-10}$, an execution time one order of magnitude faster than \textsf{starry} on CPU and up to two orders of magnitude faster on GPU for large datasets (see right plot of \autoref{fig:real_benchmark}).\\\\
\begin{figure}[H]
    \begin{center}
        \includegraphics[width=\textwidth]{../workflows/speed/figures/speed_vs_N_realistic.pdf}
        \caption{}
        \label{fig:real_benchmark}
    \end{center}
\end{figure}

@@

\section{Error budget}
\begin{figure}[H]
    \begin{center}
        \includegraphics[width=\textwidth]{../workflows/precision/figures/error_SAR.pdf}
        \caption{\codelink{workflows/precision/scripts/plot_error_SAR}}
        \label{fig:precision_SAR}
    \end{center}
\end{figure}

@@

>>>>

In \autoref{fig:precision_limbdark}, we compare the precision of our limb darkened light curve model with that of the \textsf{exoplanet} Python package (which uses the same formalism as \cite{Agol2020} but limited to linear and quadratic limb darkening). The larger errors observed in this figure are due to the transformation of the polynomial limb darkening coefficients to the spherical harmonics coefficients. For example, a set of unitary polynomial limb darkening coefficients of order 20 leads to spherical harmonics coefficients as large as $10^6$. We note that these errors scale with the values of $\bvec{u}$.\\\\

@@

As described in \autoref{solution_vector}, we compute occultation light curves using the Gauss Legendre quadrature to approximate the $\mathcal{P}$ integral involved in the expression of the solution vector $\bvec{s}^\mathsf{T}$. Hence, the precision of the computed flux $f$ depends on the order of the Legendre polynomial which defines the number of points used to approximate $\mathcal{P}$ (see \autoref{fig:relative_error_1}).Hence, users must carefully set the order parameter to balance precision and computational speed (see \autoref{speed}).

@@

With the results of this section, we validate the precision of \textsf{jaxoplanet} and show it is on per with the legacy C++ \textsf{starry} implentation\footnote{version 1.2.0 \citep{starry_120}} described in \cite{starry}.\\\\

@@

One of the advantage of the \textsf{JAX} library is its ability to perform hardware-accelerated computations on CPUs and GPUs. In this section, we evaluate the speed of the \textsf{jaxoplanet} implementation and compare it with the C++ implementation of \textsf{starry}. For quadratically limb darkened light curves, we also compare our implementation with the one provided by the \textsf{exoplanet} Python package.\\\\
While performing numerical integration may seem as the downside of our implementation, adapting the value of $q$ to the precision required for specific applications allows users to optimize the computation time of the model to their advantage. To demonstrate this point, we first consider the transmission spectroscopy of TRAPPIST-1 b performed with the JWST/NIRISS instrument (GO-2589, \citealt{Lim2023}). These observations reached an estimated precision of 89 ppm for a planet with a relative radius of the order $r\sim0.1$. In this context, numerical integration with an order as low as $q=10$ guarantees that our model reaches a precision of $10^{-16}$, an execution time on par with \textsf{exoplanet} on CPU and up to an order of magnitude faster on GPU for large datasets (see left plot of \autoref{fig:real_benchmark}). The second example we consider is the secondary ecclipse mapping of HD 189733b ($r\sim0.3$) using Spitzer's 8 micron channel \citep{majeau2012} obtained at a precision of about 1 ppt. For a spherical harmonics map of degree $l_{max}=3$ (although $l_{max}=1$ is enough in this case), $q=10$ guarantees a model with a precision of $10^{-10}$, an execution time one order of magnitude faster than \textsf{starry} on CPU and up to two orders of magnitude faster on GPU for large datasets (see right plot of \autoref{fig:real_benchmark}).\\\\
\begin{figure}[H]
    \begin{center}
        \includegraphics[width=\textwidth]{../workflows/speed/figures/speed_vs_N_realistic.pdf}
        \caption{}
        \label{fig:real_benchmark}
    \end{center}
\end{figure}

In general, we show that models computed at low values of $q$, which are very fast, match the precision of measurements from modern instruments. Although this might not be required, \autoref{precise_speed} shows reference speed benchmarks for models evaluated at state-of-the-art precision.\\\\
In order to help users set the value of $q$ according to their required precision, the \textsf{jaxoplanet} Python package provide helper functions to estimate the precision of the light curve model given the relative radius $r$, the spherical harmonics degree $l_{max}$ (if general non-uniform maps are required), and the order of the polynomial limb darkening law\footnote{See tutorial at \href{}{}}.\\\\

@@

\section{Precision}\label{precise_precision}
\begin{figure}[H]
    \begin{center}
        \includegraphics[width=\textwidth]{../workflows/precision/figures/error_SAR.pdf}
        \caption{Typical errors in the matrices used to compute the flux (\autoref{eq:starry}) with \textsf{starry} and \textsf{jaxoplanet}. \codelink{workflows/precision/scripts/plot_error_SAR}}
        \label{fig:precision_SAR}
    \end{center}
\end{figure}

@@

\begin{figure}[H]
    \begin{center}
        \includegraphics[width=\textwidth]{../workflows/precision/figures/error_order_degree.pdf}
        \caption{Maximum errors in the overall flux $f$ for all degrees of the spherical harmonics components for different occultor radii and orders $q$ of the Gauss-Legendre quadrature used to compute the P integral (see \autoref{solution_vector}). Errors shown in this figure correspond to the maximum values of the scaled errors shown in \autoref{fig:precision_s}, maxima taken over the range of values $b$.}
        \label{fig:precision_order_gausslegendre_degree}
    \end{center}
\end{figure}

\section{Speed}\label{precise_speed}
In \autoref{fig:speed_order_gausslegendre}, we show the evolution of the time required to compute the occultation light curve of a limb darkened star and that of a non-uniform star described by spherical harmonics of degree $l_{max}=20$, depending on the order $q$ of the numerical integration. On CPU, our benchmark is done on the single core of an Apple M2 max chip, while on GPU we use the single core of an NVIDIA Tesla V100 processor.\\\\
\begin{figure}[H]
    \begin{center}
        \includegraphics[width=\textwidth]{../workflows/speed/figures/speed_vs_order.pdf}
        \caption{Computation time of the occultation light curve of a limb-darkened star (left) and a non-uniform star described by spherical harmonics up to order $l_{max}=20$ (right) depending on the order $q$ of the Gauss-Legendre quadrature used to compute the $\mathcal{P}$ integral numerically (see \autoref{solution_vector}). The processing time reported for \textsf{exoplanet} and \textsf{starry} is independent of $q$ as these two codes provide closed-form solutions for the integral $\mathcal{P}$. Light curves are computed for $10\,000$ points in transit and an occultor radius $r=0.1$.}
        \label{fig:speed_order_gausslegendre}
    \end{center}
\end{figure}
Finally, \autoref{fig:speed_N} shows the evaluation time of \textsf{jaxoplanet} as a function of the number of points in the light curve compared to \textsf{exoplanet}, for a quadratically limb darkened star, and \textsf{starry}, for a surface map of degree $l_{max} = 20$.
\begin{figure}[H]
    \begin{center}
        \includegraphics[width=\textwidth]{../workflows/speed/figures/speed_vs_N.pdf}
        \caption{Computation time of the occultation light curve of a limb-darkened star (left) and a non-uniform star described by spherical harmonics up to order $l_{max}=20$ (right) depending on the number of points in the light curve. Light curves are computed at $q=200$ and an occultor radius $r=0.1$.}
        \label{fig:speed_N}
    \end{center}
\end{figure}
These benchmarks are produced for values of $q$ which guarantee that our models reach the same precision as state-of-the-art codes. As discussed in \autoref{discussion}, \textsf{jaxoplanet} can be further optimized by reducing the order $q$ of numerical integration while keeping the precision required for specific applications.
